A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-07-23 text/discussion.Rmd

One of the biggest technical hurdles in tumor research is that the search space
comprising ~20,000 protein-coding genes is comparable with the number of tumor
samples, even with multiple mutations per sample
[@hornNetSigNetworkbasedDiscovery2018]. And that is not accounting for the fact
that ~30k non-coding genes and hundreds of thousands of non-coding regulatory
elements exist with oncogenic or tumor suppressor potential in spite of many
efforts detect non-random or selected mutation above the noise[@deesMuSiC2012;
@tamboreroOncodriveCLUST2013; @lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @jiangWITER2019;
@zhaoModelingPositive2019; @non-coding-mutations-Zhang]. Hence, we sought to
investigate cancer as a disease of basic cellular and biochemical pathways,
which we accomplished by translating gene-level mutations into pathway level
disruptions.

The hypothesis that cancer is the result of dysfunction from a limited number of
basic cellular processes common to all living human cells was first
introduced and later expanded on in a pair of classic essays by Hanahan and
Weinberg [@weinberg200; @hallmarks_II]. In this view all tumors, regardless of
tissue of origin, acquire 6 essential phenotypes; sustaining proliferative
signaling, evading growth suppressors, replicative immortality, evading
apoptosis, supportive angiogenesis, and tissue invasion and metastasis
[@weinberg2000]. Advancements in the field over the ensuing decade led to the
inclusion of new tumor promoting processes, namely deregulation of celullar
energy cycle, tumor promoting inflammation, evasion of the immune system, genome
instability, and on top of these acquired traits a new appreciation for the
concept of the tumor microenvironment and its contribution to the evolution and
promotion of metastatic tumorigenesis [@hallmarks_II].

An alternative hypothesis is that every tumor belongs to one of a large number
of syndromes which are unique to each tissue-of-origin, that share some
mechanisms and treatment strategies. Recent publication of Encode, Roadmap and TCGA
consortium papers present a view largely consistent with the latter
hypothesis. Perhaps owing to the intractable complexity of genomic data in all
its forms, the inescapable conclusion thus far is that tissue-of-origin remains
the most important driver of tumor characteristics at every scale and by every
technique. Our observations contrast with this view, and instead support an
interpretation of publicly available  data in which all tumors manifest one of a
limited number of metastable phenotypes resulting from disruptions in basic
pathways represented by our classes (**figure 1a-k**), with increased fitness
regardless of tissue of origin, which then converge on end stage cancer by
accumulating complementary mutations to the other classes (**figure 1l** and
**Table 2**).

>>>>

One of the biggest hurdles in cancer research is that the search space
comprising ~20,000 protein-coding genes is comparable with the number of tumor
samples, even with multiple mutations per sample. And that is not accounting for
the fact that ~30k non-coding genes and hundreds of thousands of non-coding
regulatory elements exist with oncogenic or tumor suppressor potential
[@deesMuSiC2012;
@tamboreroOncodriveCLUST2013; @lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @jiangWITER2019;
@zhaoModelingPositive2019; @non-coding-mutations-Zhang]. Hence, we sought to
investigate cancer as a disease of basic cellular and biochemical pathways,
which we accomplished by translating gene-level mutations into pathway level
disruptions. Our approach differs from previously proposed methods [@PATHNET] in
that we chose to focus on pathways as the unit of disruption instead of the
gene, using the conceit that individual mutations may be sufficient to disrupt
entire pathways (as is often the case with drivers). Rather than evaluating the
burden of mutations on each pathway within or across tumors, which can be biased
by the number of genes or the coverage of pathway-linked genes in the genome,
and which we find to be relatively uniformly distributed, we scored each pathway
in binary fashion, allowing one mutation or multiple mutations to count equally
in a sample. To our knowledge, this approach has not previously been attempted
in the literature in spite of its relative simplicity.

We limited our analysis to mutations with likely deleterious effects, including
missense, nonsense, frame-shift stop-gain and splice site mutations in genes
that are actively expressed in each cancer type, thus avoiding bias from
transcription coupled repair. Our method of filtration differs from the
"rank-and-cut" method of [@PCAWG-MAIN] but represents a reasonable attempt to
account for the same biases. We also filtered genes on the basis of pathway
membership in biochemical pathways, excluding curated genesets related to
diseases, syndromes, or classes of proteins ( _e.g._ transcription factors). We
chose this approach to limit redundancy, which is already high in the remaining
pathways for some driver genes, and to exclude biologically unrelated
collections of genes. We do not account for non-coding mutations, amplification
events or structural variants, though we hope to include them in subsequent
studies (see below). Our pathway-centric analysis also allows us to circumvent
the problem that some cancer-promoting mutations likely occur at low enough
frequencies as to be indistinguishable from background. Attempts to capture
these associations better using Bayesian networks have yielded important new
insights into this problem and new genes that would not have been possible with
traditional analysis [@hornNetSigNetworkbasedDiscovery2018], supporting the case
that pathway-centric approaches like ours have the potential to increase
statistical power.

The hypothesis that cancer is the result of dysfunction from a limited number of
basic cellular processes common to all living human cells was first
introduced and later expanded on in a pair of classic essays by Hanahan and
Weinberg [@weinberg200; @hallmarks_II]. In this view, all tumors regardless of
tissue of origin acquire 6 essential phenotypes; sustaining proliferative
signaling, evading growth suppressors, replicative immortality, evading
apoptosis, supportive angiogenesis, tissue invasion and metastasis
[@weinberg2000]. Advancements in the field over the ensuing decade led to the
inclusion of new tumor promoting processes, namely deregulation of cellular
energy cycle, tumor promoting inflammation, evasion of the immune system, genome
instability, and on top of these acquired traits a new appreciation for the
concept of the tumor microenvironment and its contribution to the evolution and
promotion of metastatic tumorigenesis [@hallmarks_II].

An alternative hypothesis is that every tumor belongs to one of a large number
of syndromes which are unique to each tissue-of-origin, that share some
mechanisms and treatment strategies. Recent publication of TCGA consortium
papers present a view largely, and surprisingly, consistent with this hypothesis
[@hoadleyMultiplatformAnalysis2014; @hoadleyCellofOriginPatternsDominate2018].
Perhaps owing to the intractable complexity of genomic, proteomic data and
patient metadata in all its forms, the inescapable conclusion thus far is that
tissue-of-origin remains the most important driver of tumor characteristics at
every scale and by every measure. Our observations contrast with this view, and
instead support an interpretation of publicly available data in which all tumors
manifest one of a limited number of metastable phenotypes resulting from
disruptions in basic pathways represented by our classes (**figure 1a-k**). Each
of these classes may represent unique combinations of pathway disruptions with
increased fitness regardless of tissue of origin, which then converge on end
stage cancer by accumulating complementary mutations to the other classes
(**figure 1l** and **Table 2**).
